Linear Reduction: Parsing and Interpreting Code Without Trees

### 2024-11-05 Paper/paper.tex

%\usepackage[margin=1in]{geometry}
\usepackage{amsmath, amssymb, mathtools}
\usepackage{enumitem}
\usepackage{hyperref}
\usepackage{float}

@@

\begin{algo}\label{alg:eval}
\caption{\eval uating arithmetical expressions}
\algorithm
\Function{eval}{$\xi$}
    \If{$\xi=\s x\in\s X$}
        \State\Return $\s x$
    \ElseIf{$\xi[0]=\s L$}
        \State $\xi$ is of the form $\s L\rho\s R\xi'$
        \State\Return $\eval(\eval(\rho)\xi')$
    \Else
        \Comment Find the greatest index with the smallest operator.\cr
        Note that $\max_i{\rm argmin}_if(i)$ is the maximal $i$ which minimizes $f$.\EndComment
        \State $i\coloneqq\max_i{\rm argmin}_i\set{\pi(\xi[i])}[{\xi[i]\in\s S,\unparen(\xi,i)}]$
        \State $\s s\coloneqq\xi[i]$
        \Comment $\xi'=\xi[:i-1]$, $\xi''=\xi[i+1:]$ \EndComment
        \State $\xi$ is of the form $\xi'\s s\xi''$
        \State\Return $f_{\s s}(\eval(\xi'),\eval(\xi''))$
    \EndIf
\EndFunc
\ealgorithm
\end{algo}

@@

>>>>

%\usepackage[margin=1in]{geometry}
\usepackage{amsmath, amssymb, mathtools}
\usepackage{enumitem}
\usepackage{float}
\usepackage{hyperref}
\usepackage{listings}
\usepackage{graphicx}

@@

\begin{algo}
\algorithm
\Function{eval}{$\xi$}
    \If{$\xi=\s x\in\s X$}
        \State\Return $\s x$
    \ElseIf{$\xi[0]=\s L$}
        \State $\xi$ is of the form $\s L\xi'\s R\zeta$
        \State\Return $\eval(\eval(\xi')\zeta)$
    \Else
        \Comment Find the greatest index with the smallest operator.\cr
        Note that $\max_i{\rm argmin}_if(i)$ is the maximal $i$ which minimizes $f$.\EndComment
        \State $i\coloneqq\max_i{\rm argmin}_i\set{\pi(\xi[i])}[{\xi[i]\in\s S,\unparen(\xi,i)}]$
        \State $\s s\coloneqq\xi[i]$
        \Comment $\xi'=\xi[:i-1]$, $\xi''=\xi[i+1:]$ \EndComment
        \State $\xi$ is of the form $\xi'\s s\xi''$
        \State\Return $f_{\s s}(\eval(\xi'),\eval(\xi''))$
    \EndIf
\EndFunc
\ealgorithm
\caption{\eval uating arithmetical expressions}
\label{alg:eval}
\end{algo}

@@

\section{Comparing Implementations}

In this section we will comapre the runtimes of both implementations (one using linear reduction and one using menhir\NEEDCITE{}).
A few notes beforehand:
\begin{itemize}
    \item The linear reduction implementation has more features than the menhir implementation, this should not severely impact the comparisons and we will of course not be comparing the implementations
            on code which the menhir implementation does not support.
    \item LLang supports left composition.
            Meaning code of the form \ttt{f x y} is read as \ttt{(f x) y}.
            We found no way of supporting this with menhir.
            This is because our grammar for the menhir implementation contains the following fragment:
            \[ {\it value} \ccoloneqq {\it const}\gbar{\it ident}\gbar{\it funcall},\qquad {\it funcall}\ccoloneqq{\it value}\ {\it value} \]
            So \ttt{f x y} can be parsed as \ttt{f (x y)} instead:

            \centerline{\begin{tikzpicture}[
                    level 1/.style = {sibling distance=10em},
                    level 2/.style = {sibling distance=10em},
                    level 3/.style = {sibling distance=10em},
                    level 4/.style = {sibling distance=10em},
                    level 5/.style = {sibling distance=10em},
                    level 6/.style = {sibling distance=10em},
                    ]
                \node {\it value}
                    child {node {\it funcall}
                        child {node {\it value}
                            child {node {\it ident}
                                child {node {\ttt f}}
                            }
                        }
                        child {node {\it value}
                            child {node {\it funcall}
                                child {node {\it value}
                                    child {node {\it ident}
                                        child {node {\ttt x}}
                                    }
                                }
                            }
                            child {node {\it funcall}
                                child {node {\it value}
                                    child {node {\it ident}
                                        child {node {\ttt y}}
                                    }
                                }
                            }
                        }
                    };
            \end{tikzpicture}
            }
\end{itemize}

\subsection{Currying}

We begin with a simple program to demonstrate currying.
Currying is the action of taking a function $f\colon X\times Y\longto Z$ and converting it to a function $c(f)\colon X\longto(Y\longto Z)$ where $c(f)(x)\colon y\mapsto f(x,y)$.
Our LLang code is as follows:

\lstinputlisting[language=Caml, frame=single]{../Demos/currying.ml}

Comparing the time it takes both implementations to run this 300 times, we see that it takes linear reduction 0.0040886 seconds, opposed to the classical implementation which takes 0.0036926 seconds.

\subsection{Arithmetical Expressions}

We can generate random arithmetical expressions using the following algorithm:

\medskip
\algorithm
\Comment Create an expression with $\ell$ subterms, with a parentheses depth of $d$,\cr where numbers are chosen from $[0,n)$.\EndComment
\Function{Generate-Expression}{$\ell,d,n$}
    \lIf{$d=0$} \vtop{\hsize=.8\hsize
        \noindent \Return an expression with $\ell$ randomly chosen numbers in\hfil\break $[0,n)$ and $\ell-1$ random operators in $\set{+,-,\times,\div}$.}
    \State \Return $\bigl(\hbox{\tencsc Generate-Expression}(\ell,d-1,n)\bigr)\circ_1\cdots\circ_{\ell-1}$
    \nonum\State \hfill$\bigl(\hbox{\tencsc Generate-Expression}(\ell,d-1,n)\bigr)$
    \nonum\State where $\circ_i\in\set{+,-,\times,\div}$ are chosen randomly
\EndFunc
\ealgorithm
\medskip

\noindent We then run LLang code of the form \ttt{let x = <expr>;} where \ttt{<expr>} is generated by the above algorithm.
We measure how long it takes to run this code five times with expressions generated by $\hbox{\tencsc Generate-Expression}(n,n,100)$ as $n$ varies.
Doing so produces the following graph in figure \ref{fig:expr}.

\begin{figure}

    \centerline{\includegraphics[width=\hsize]{../Docs/Images/expr.png}}
    \caption{Time to compute $\hbox{\tencsc Generate-Expression}(n,n)$}
    \label{fig:expr}

\end{figure}

\subsection{Fibonacci}

The fibonacci sequence is a linear recurrence defined by $F_n=F_{n-1}+F_{n-2}$ for $n>1$ and $F_1=F_0=1$.
We wrote a script to compute the $n$th fibonacci number recursively in LLang:

\begin{lstlisting}[language=Caml, frame=single]
fun fib (x) {
    if (x < 2) {
        1
    }{
        (fib (x-1)) + (fib (x-2))
    }
}
let x = (fib n);
_prim_print x
\end{lstlisting}
